---
title: "DRY code examples"
format:
  html:
    theme: spacelab
    toc: true
    number-sections: true
    author: Sara Correa García
---

# Load libraries
```{r}
pacman::p_load(pacman, broom, here, doBy, tidyverse)
```

# Creating summary tables with doBy

## Load data
```{r}
prok.treat = read.csv(file = here("prok.treat.csv"), row.names = 1)
```

## Inspect the data
```{r}
dim(prok.treat)
str(prok.treat)
View(prok.treat)
```

## Summarize data
You can look at basic summary statistics one by one
```{r}
adhP.sum = summary(prok.treat$adhP)
adhP.sum
aliA.sum = summary(prok.treat$aliA)
aliA.sum
badH.sum = summary(prok.treat$badH)
badH.sum
bamB.sum = summary(prok.treat$bamB)
bamB.sum
# This can quickly become too repetitive. Also, what if I had 1000 instead of 16 genes?
```

We can dry our code by looking at paterns that repeat in our code that we can abstract.

## Evaluate Summary statistics for all gene groups using specific libraries
```{r}
library(doBy)
colnames(prok.treat)
genes_summary = as_tibble(summaryBy(adhP + aliA + badH + bamB + bamD + fadJ + fadN + gabD + mhpD + mhpE + phaZ + pobA + ppk + putA + soxY + soxZ ~ group, data = prok.treat, FUN = function(x) { c(mean = mean(x), sd = sd(x)) } ) ) |> 
    mutate(across(where(is.numeric), ~ num(., digits = 2)))
write_csv(genes_summary, file = here( "genes-summary-counts-cpm.csv"))
```


## Alternatively a drier version

```{r}
# Select only numerical variables (NA degradation and absolute abundance of genes, and the grouping variable)
prok.treat.dry = prok.treat |> 
               select_at(vars(group,names(prok.treat)[map_lgl(prok.treat, is.numeric)])) 
# Create a vector genes_to-test pasting the column names of our genes (starting at 3, because column 1 is group and column 2 is NA_ppm and going all the way up to the last column). The argument collapse indicates the separation between colnames. We include a + because we want a string that can be passed as a formula later on.
genes_to_test = paste(c(names(prok.treat.dry[,3:ncol(prok.treat.dry)])), collapse = " + ") 
#Use the function doBy::summaryBy coupled with the function as.formula to calculate the mean and the sd for all the genes in the string. 
genes_summary_dry = as_tibble(summaryBy(as.formula(paste(genes_to_test, "~ group")), data = prok.treat.dry, FUN = function(x) { c(mean = mean(x), sd = sd(x))})) |> 
    mutate(across(where(is.numeric), ~ num(., digits = 2)))
```




# Kruskal Wallis tests. Looping!

Using the same data set, calculate a kruskal Wallis test for all genes. 

In this exercise, you will need to complete the code. There are missing parts (function, objects...) replaced by and "#". 

--- Let's start!
You first need to transform as a tibble. Then, you need to gather the genes that belong to the same group level. You can use the tidyverse to make it easier. 

```{r}
dry.kw = as_#(data.frame(prok.treat.dry)) |> #(group)
```

--- Too difficult? You can take a look
```{r}
dry.kw = as_tibble(data.frame(prok.treat.dry)) |> group_by(group)
```

--- 
Use gather to pair key-value (gene name-abundance) pairs and pivot to long format. Remember, gather requires data to be as data.frame, so you need to previously transform dry.kw to fit!
Then, you can group by key (gene name). Finally, you can run the kruskal wallis
```{r}
dry.kw = ##.data.frame(dry.kw)
dry.kw.loop = dry.kw |> gather(key, value, -####)  |> # Don't forget to exclude the group variable
  group_by(###) |>  # Group by gene name
  do(tidy(kruskal.test(x = .$###, g = .$###))) # To know more about which arguments are passing to the kruskal.test function you can type ?kruskal.test in the console. 
```

--- Hint

```{r}
dry.kw = as.data.frame(dry.kw)
dry.kw.loop = dry.kw |> gather(key, value, -group)  |>
  group_by(key) |>  
  do(tidy(kruskal.test(x = .$value, g = .$group)))
```

--- Moving on to add the cherry on top

Because we calculated a large number of test, the chances to find significant values when in reality they are not, have escalated as quickly as our performance when we apply dry code. 
To avoid it, we can adjust our p values to reduce our false discovery rate.
Keeping up with the tibble, we can pass the fucntion mutate to create a new variable in our dry.kw.loop.
Alternatively, we can do it after transforming our table into a dataframe. 
Finally, we can save our results using the here() package
```{r}
dry.kw.loop |> 
   mutate(pval.adj = ###(p.value, method = "BH")) # What is the name of the function
as.data.frame(dry.kw.loop)
dry.kw.loop$### = ####(dry.kw.loop$p.value, method = "##") # select a method of your choice
write.csv(dry.kw.loop, file = #(#gene_KW.csv#))
```

--- Hint

```{r}
dry.kw.loop |> 
   mutate(pval.adj = p.adjust(p.value, method = "BH"))# check
as.data.frame(dry.kw.loop)
dry.kw.loop$p.adjust = p.adjust(dry.kw.loop$p.value, method = "BH")
write.csv(dry.kw.loop, file = here("gene_KW.csv"))
```



--- Take a look at the final code

```{r}
prok.treat.dry
dry.kw = as_tibble(data.frame(prok.treat.dry)) |> group_by(group) #
dry.kw = as.data.frame(dry.kw)
dry.kw.loop = dry.kw |> gather(key, value, -group)  |>
  group_by(key) |>  
  do(tidy(kruskal.test(x = .$value, g = .$group)))
dry.kw.loop |> 
   mutate(pval.adj = p.adjust(p.value, method = "BH"))# check
as.data.frame(dry.kw.loop)
dry.kw.loop$p.adjust = p.adjust(dry.kw.loop$p.value, method = "BH")
write.csv(dry.kw.loop, file = here("gene_KW.csv"))
```

